---
title: "badri analysis"
author: "Merly Klaas"
date: "11/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lavaan)
```

```{r}
lower <- '
6.964
4.399 6.361
2.691 2.576 6.647
1.191 1.080 0.960 1.339
1.405 1.266 1.067 0.821 1.573
1.827 1.398 1.122 0.821 0.971 2.804
1.698 1.319 1.060 0.771 0.940 2.221 2.350
1.656 1.279 1.073 0.753 0.900 2.197 2.044 2.393
1.675 1.302 1.075 0.806 0.966 1.974 1.952 1.900 2.826
1.540 1.219 0.953 0.719 0.874 1.968 1.808 1.825 1.868 2.480
1.629 1.248 1.030 0.744 0.903 2.224 2.033 2.080 1.956 1.999 2.804
1.771 1.365 1.084 0.762 0.926 2.224 2.056 2.060 1.930 1.905 2.220 2.802
0.530 0.491 0.383 0.252 0.262 0.280 0.256 0.259 0.282 0.268 0.257 0.257 0.482
0.574 0.559 0.426 0.273 0.247 0.236 0.229 0.234 0.240 0.221 0.222 0.239 0.214 0.532
0.302 0.269 0.200 0.176 0.174 0.165 0.168 0.163 0.181 0.161 0.163 0.166 0.162 0.132 0.334

'

badri.cov <-
  getCov(lower,
         names = c("Q9", "Q10", "Q11", "Q17_1", "Q17_19", "Q18_1", "Q18_2", "Q18_3", "Q18_4", "Q18_5", "Q18_6", "Q18_7", "Q13", "Q14", "Q15"))

```

## Model 1 Specifying measurement model

```{r}

badri_cfa_1 <-'
#specifying measurement model portion
  home =~ Q13 + Q14 + Q15
  scho =~ Q18_1 + Q18_2 + Q18_3 + Q18_4 + Q18_5 + Q18_6 + Q18_7 
  happy =~ Q9 + Q10 + Q17_1 + Q11 + Q17_19
'



badri_cfa_fitted_1 <- cfa(badri_cfa_1, 
                        sample.cov = badri.cov, 
                        sample.nobs = 14828)
summary(badri_cfa_fitted_1, standardized = TRUE,
        fit.measures = TRUE)

modindices(badri_cfa_fitted_1, sort= TRUE)
```

## Model 2 adding structural path
  * According to the paper: The steps involved the inclusion of the error covariance between variables, adding paths between latent constructs, and checking for the largest standardized residuals.

```{r}
badri_cfa_2 <-'
#specifying measurement model portion
  home =~ Q13 + Q14 + Q15
  scho =~ Q18_1 + Q18_2 + Q18_3 + Q18_4 + Q18_5 + Q18_6 + Q18_7 
  happy =~ Q9 + Q10 +  Q17_1 + Q11 + Q17_19
  
#specifying structural model portion  
happy ~ scho + home
scho ~ home

#trying to add covariances but it intoroduced error:  lavaan ERROR: could not compute modification indices; information matrix is singular
# Q9 ~~ Q10 + Q17_1 + Q11 + Q17_19
# Q10 ~~ Q17_1 + Q11 + Q17_19
# Q17_1 ~~ Q11 + Q17_19
# Q11 ~~ Q17_19
# 
# 
# Q18_1 ~~ Q18_2 + Q18_3 + Q18_4 + Q18_5 + Q18_6 + Q18_7
# Q18_2 ~~ Q18_3 + Q18_4 + Q18_5 + Q18_6 + Q18_7
# Q18_3 ~~ Q18_4 + Q18_5 + Q18_6 + Q18_7
# Q18_4 ~~ Q18_5 + Q18_6 + Q18_7
# Q18_5 ~~ Q18_6 + Q18_7
# Q18_6 ~~ Q18_7
# 
# Q13 ~~ Q14 + Q15
# Q14 ~~ Q15

'



badri_cfa_fitted_2 <- cfa(badri_cfa_2, 
                        sample.cov = badri.cov, 
                        sample.nobs = 14828)
summary(badri_cfa_fitted_2, standardized = TRUE,
        fit.measures = TRUE)

fit_stats <- fitmeasures(badri_cfa_fitted_2)
modindices(badri_cfa_fitted_2, sort= TRUE)


```

## Model 3 modification based on modification indices



```{r}
badri_cfa_3 <-'
#specifying measurement model portion
  home =~ Q13 + Q14 + Q15
  scho =~ Q18_1 + Q18_2 + Q18_3 + Q18_4 + Q18_5 + Q18_6 + Q18_7 
  happy =~ Q9 + Q10 + Q17_1 + Q11 + Q17_19
  
#specifying structural model portion  
happy ~ scho + home
scho ~ home

#covariances 

Q9 ~~ Q10 
Q17_1 ~~ Q17_19
Q9 ~~ Q17_1
Q9 ~~ Q17_19
Q10 ~~ Q17_1
Q10 ~~ Q17_19

'



badri_cfa_fitted_3 <- cfa(badri_cfa_3, 
                        sample.cov = badri.cov, 
                        sample.nobs = 14828)
summary(badri_cfa_fitted_3, standardized = TRUE,
        fit.measures = TRUE)

fit_stats <- fitmeasures(badri_cfa_fitted_3)
modindices(badri_cfa_fitted_3, sort= TRUE)

```

## Model 4 Modification : Experience at school is impacted by students' experience at home and happiness

```{r}
badri_cfa_4 <-'
#specifying measurement model portion
  home =~ Q13 + Q14 + Q15
  scho =~ Q18_1 + Q18_2 + Q18_3 + Q18_4 + Q18_5 + Q18_6 + Q18_7 
  happy =~ Q9 + Q10 + Q17_1 + Q11 + Q17_19
  
#specifying structural model portion  
scho ~ happy + home
happy ~ home

#covariances 
Q9 ~~ Q10 
Q17_1 ~~ Q17_19
Q9 ~~ Q17_1
Q9 ~~ Q17_19
Q10 ~~ Q17_1
Q10 ~~ Q17_19

'



badri_cfa_fitted_4 <- cfa(badri_cfa_4, 
                        sample.cov = badri.cov, 
                        sample.nobs = 14828)
summary(badri_cfa_fitted_4, standardized = TRUE,
        fit.measures = TRUE)

fit_stats <- fitmeasures(badri_cfa_fitted_4)


```

